What Should I Test?

### 2024-02-07 mss.tex
Rework some listings.

Here's an extreme unit test, some may even say it is not a unit test, but an integration test (or something else).

First, we need to build out a little more infrastructure, because we are doing pokes on the Obelisk database engine.

\begin{lstlisting}
::
::  Build an example bowl manually
++  bowl
  |=  [run=@ud now=@da]
  ^-  bowl:gall
  :*  [~zod ~zod %obelisk `path`(limo `path`/test-agent)]  :: (our src dap sap)
      [~ ~ ~]                                              :: (wex sup sky)
      [run `@uvJ`(shax run) now [~zod %base ud+run]]       :: (act eny now byk)
  ==
\end{lstlisting}

\noindent
Now we are ready to test the evolution through time of a database, as in Listing~\ref{lst:evolution}.

\begin{lstlisting}[
  float,
  style=listingcode,
  caption={Sequent tests},
  label={lst:evolution}
]e
::
::  time, insert as of 1 second > schema
++  test-time-insert-gt-schema
  =|  run=@ud
  =^  mov1  agent  
    %:  ~(on-poke agent (bowl [run ~2000.1.1]))
        %obelisk-action
        !>([%tape-create-db "CREATE DATABASE db1"])
    ==
  =.  run  +(run)
  =^  mov2  agent  
    %:  ~(on-poke agent (bowl [run ~2000.1.2]))
        %obelisk-action
        !>  :+  %tape
                %db1
                "CREATE TABLE db1..my-table (col1 @t) PRIMARY KEY (col1) ".
                "AS OF ~2023.7.9..22.35.35..7e90"
    ==
  =.  run  +(run)
  =^  mov3  agent  
    %:  ~(on-poke agent (bowl [run ~2023.7.9..22.35.35..7e90]))
        %obelisk-action
        !>  :+  %tape 
                %db1 
                "INSERT INTO db1..my-table (col1) VALUES ('cord') ".
                "AS OF ~2023.7.9..22.35.36..7e90"
    ==
  =+  !<(=state on-save:agent)
  ;:  weld
  %+  expect-eq
    !>  :-  %results
          :~  [%result-da 'data time' ~2023.7.9..22.35.36..7e90]
              [%result-ud 'row count' 1]
              ==
    !>  ->+>+>.mov3
  %+  expect-eq
    !>  db-time-insert-tbl
    !>  databases.state
  ==
\end{lstlisting}

@@

\begin{lstlisting}[
  float,
  style=listingcode,
  caption={Multiple tests in one bundle},
  label={lst:bundle}
]
::
:: common things to test
:: 1) basic command works producing AST object
:: 2) multiple ASTs
:: 3) all keywords are case ambivalent
:: 4) all names follow rules for faces
:: 5) all qualifier combinations work
::
:: alter index
::
:: tests 1, 2, 3, 5
::       extra whitespace characters
::       multiple command script:
::         alter index... db.ns.index db.ns.table columns action %disable
::         alter index db..index db..table one column action %rebuild
++  test-alter-index-1
  =/  expected1
    :*  %alter-index
        [%qualified-object ship=~ database='db' namespace='ns' name='my-index'] 
        [%qualified-object ship=~ database='db' namespace='ns' name='table']
        :~  [%ordered-column name='col1' is-ascending=%.y]
            [%ordered-column name='col2' is-ascending=%.n]
            [%ordered-column name='col3' is-ascending=%.y]
            ==
        %disable
        ==
  =/  expected2
    :*  %alter-index
        [%qualified-object ship=~ database='db' namespace='dbo' name='my-index']
        [%qualified-object ship=~ database='db' namespace='dbo' name='table']
        ~[[%ordered-column name='col1' is-ascending=%.y]]
        %rebuild
        ==
  %+  expect-eq
    !>  ~[expected1 expected2]
    !>  %-  parse:parse(default-database 'db1')
      "aLter \0d INdEX\09db.ns.my-index On db.ns.table  ".
      "( col1  asc , col2\0a desc  , col3) \0a dIsable \0a;\0a aLter \0d ".
      "INdEX\09db..my-index On db..table  ( col1  asc ) \0a \0a rEBuild "
\end{lstlisting}

@@

Peano arithmetic is a better lay programmer's introduction to proof by induction. There exists (or perhaps \emph{does not exist}) a special concept, zero. Playing fast and loose with classical logic, “ex nihil sequitur quodlibet”, from nothing (more commonly a falsehood, or a contradiction) follows everything (i.e. anything).\footnote{The “principle of explosion” rigorously follows from use of disjunctive syllogisms.  Here, we jocosely indicate that by proceeding from zero inductively we can demonstrate desired properties of testing.} Zero is not only an integer; rather, in terms of Nock-based programming, \lstinline[style=inlinecode]{~} is not only the beginning of counting,\footnote{Yarvin opted to invert \texttt{true} and \texttt{false} (e.g. \texttt{urbit/vere} commit \texttt{a8c1a799}, 2014-11-04), meaning that Nock loobeans do not align inductively with other inductive types.} but the nothing of every inductive type, most importantly trees and lists. When speaking to programmers, Peano tells us there is a universal nothing \lstinline[style=inlinecode]{~} and there is a function called successor $\textrm{Succ}$, which produces some next thing from a previous thing (or lack of thing). Applying $\textrm{Succ}$ to nothing gives us the first thing.  Applying $\textrm{Succ}$ to the first thing gives us the second thing, and so on.

@@

\begin{lstlisting}  
~|("cannot add duplicate key: {<row-key>}" !!)
\end{lstlisting}

@@

\begin{lstlisting}[
  float,
  style=listingcode,
  caption={Testing for expected error message},
  label={lst:error}
]
::
::  +expect-fail-message
++  expect-fail-message
  |=  [msg=@t a=(trap)]
  ^-  tang
  =/  b  (mule a)
  ?-  -.b
    %|  |^
        =/  =tang  (flatten +.b)
        ?:  ?=(^ (find (trip msg) tang))
          ~
        ['expected error message - not found' ~]
        ++  flatten
          |=  tang=(list tank)
          =|  res=tape
          |-  ^-  tape
          ?~  tang  res
          $(tang t.tang, res (weld ~(ram re i.tang) res))
        --
    %&  ['expected failure - succeeded' ~]
  ==
\end{lstlisting}

\begin{lstlisting}[
  float,
  style=listingcode,
  caption={urQL tests},
  label={lst:urql}
]
::
:: fail on dup rows
++  test-fail-insert-dup-rows
  =|  run=@ud
  =/  my-insert
    "INSERT INTO db1..my-table-3 (col1, col2, col3)  ".
    "VALUES ('cord',~zod,20) ('Default',Default, 0)"
  %+  expect-fail-message
    'cannot add duplicate key:'
  |.  %-  process-cmds
      :+  gen3-dbs  :: <- one key already exists
        (bowl [run ~2031.1.1])
      (parse:parse(default-database 'db1') my-insert)
\end{lstlisting}

>>>>

Here's a more extreme unit test (which some may even prefer to call an integration test):

\begin{lstlisting}[style=listingcode]
::
::  Build an example bowl manually
++  bowl
  |=  [run=@ud now=@da]
  ^-  bowl:gall
  :*  [~zod ~zod %obelisk `path`(limo `path`/test-agent)]  :: (our src dap sap)
      [~ ~ ~]                                              :: (wex sup sky)
      [run `@uvJ`(shax run) now [~zod %base ud+run]]       :: (act eny now byk)
  ==
\end{lstlisting}

\noindent
Now we are ready to test the evolution through time of a database.

\begin{lstlisting}[style=listingcode]
::
::  time, insert as of 1 second > schema
++  test-time-insert-gt-schema
  =|  run=@ud
  =^  mov1  agent  
    %:  ~(on-poke agent (bowl [run ~2000.1.1]))
        %obelisk-action
        !>([%tape-create-db "CREATE DATABASE db1"])
    ==
  =.  run  +(run)
  =^  mov2  agent  
    %:  ~(on-poke agent (bowl [run ~2000.1.2]))
        %obelisk-action
        !>  :+  %tape
                %db1
                "CREATE TABLE db1..my-table (col1 @t) PRIMARY KEY (col1) ".
                "AS OF ~2023.7.9..22.35.35..7e90"
    ==
  =.  run  +(run)
  =^  mov3  agent  
    %:  ~(on-poke agent (bowl [run ~2023.7.9..22.35.35..7e90]))
        %obelisk-action
        !>  :+  %tape 
                %db1 
                "INSERT INTO db1..my-table (col1) VALUES ('cord') ".
                "AS OF ~2023.7.9..22.35.36..7e90"
    ==
  =+  !<(=state on-save:agent)
  ;:  weld
  %+  expect-eq
    !>  :-  %results
          :~  [%result-da 'data time' ~2023.7.9..22.35.36..7e90]
              [%result-ud 'row count' 1]
              ==
    !>  ->+>+>.mov3
  %+  expect-eq
    !>  db-time-insert-tbl
    !>  databases.state
  ==
\end{lstlisting}

@@

\begin{lstlisting}[
  float,
  style=listingcode,
  caption={Multiple tests in one bundle},
  escapeinside=``,
  label={lst:bundle}
]
:: common things to test
:: 1) basic command works producing AST object
:: 2) multiple ASTs
:: 3) all keywords are case ambivalent
:: 4) all names follow rules for faces
:: 5) all qualifier combinations work
::
:: alter index
::
:: tests 1, 2, 3, 5
::       extra whitespace characters
::       multiple command script:
::         alter index... db.ns.index db.ns.table columns action %disable
::         alter index db..index db..table one column
::         action %rebuild
++  test-alter-index-1
  =/  expected1
    :*  %alter-index
        :*  %qualified-object
            ship=~ database='db'
            namespace='ns' name='my-index'
        ==
        :*  %qualified-object
            ship=~ database='db'
            namespace='ns' name='table'
        ==
        :~  :*  %ordered-column
                name='col1'
                is-ascending=%.y
            ==
            :*  %ordered-column
                name='col2'
                is-ascending=%.n
            ==
            :*  %ordered-column
                name='col3'
                is-ascending=%.y
        ==  ==
        %disable
        ==
  `\newpage`
  =/  expected2
    :*  %alter-index
        :*  %qualified-object
            ship=~ database='db'
            namespace='dbo' name='my-index'
        ==
        :*  %qualified-object
            ship=~ database='db'
            namespace='dbo' name='table'
        ==
        :~  :*  %ordered-column
                name='col1'
                is-ascending=%.y
        ==  ==
        %rebuild
        ==
  %+  expect-eq
    !>  ~[expected1 expected2]
    !>  %-  parse:parse(default-database 'db1')
      "aLter \0d INdEX\09db.ns.my-index On db.ns.table  ".
      "( col1  asc , col2\0a desc  , col3) \0a dIsable \0a;\0a aLter \0d ".
      "INdEX\09db..my-index On db..table  ( col1  asc ) \0a \0a rEBuild "
\end{lstlisting}

@@

Peano arithmetic is a better lay programmer's introduction to proof by induction. There exists (or perhaps \emph{does not exist}) a special concept, zero. Playing fast and loose with classical logic, “ex nihil sequitur quodlibet”, from nothing (more commonly a falsehood, or a contradiction) follows everything (i.e. anything).\footnote{The “principle of explosion” rigorously follows from use of disjunctive syllogisms.  Here, we jocosely indicate that by proceeding from zero inductively we can demonstrate desired properties of testing.} Zero is not only an integer; rather, in terms of Nock-based programming, \lstinline[style=inlinecode]{~} is not only the beginning of counting,\footnote{Yarvin opted to invert \texttt{true} and \texttt{false} (e.g. \texttt{urbit/vere} commit \texttt{a8c1a799}, 2014-11-04), meaning that Nock loobeans do not align inductively with other inductive types.} but the nothing of every inductive type, most importantly trees and lists. When speaking to programmers, Peano tells us there is a universal nothing \lstinline[style=inlinecode]{~} and there is a function called successor $\textrm{Succ}$, which produces some next thing from a previous thing (or lack of thing). Applying $\textrm{Succ}$ to nothing, $\textrm{Succ}(0)$, gives us the first thing.  Applying $\textrm{Succ}$ to the first thing, $\textrm{Succ}(\textrm{Succ}(0))$ gives us the second thing, and so on.

@@

\begin{lstlisting}[style=listingcode]
~|("cannot add duplicate key: {<row-key>}" !!)
\end{lstlisting}

@@

\begin{lstlisting}[
  float,
  style=listingcode,
  caption={Testing for expected error message},
  label={lst:error}
]
::
::  +expect-fail-message
++  expect-fail-message
  |=  [msg=@t a=(trap)]
  ^-  tang
  =/  b  (mule a)
  ?-  -.b
    %|  |^
        =/  =tang  (flatten +.b)
        ?:  ?=(^ (find (trip msg) tang))
          ~
        ['expected error message - not found' ~]
        ++  flatten
          |=  tang=(list tank)
          =|  res=tape
          |-  ^-  tape
          ?~  tang  res
          %=  $
            tang  t.tang
            res   (weld ~(ram re i.tang) res)
          ==
        --
    %&  ['expected failure - succeeded' ~]
  ==
\end{lstlisting}

\begin{lstlisting}[
  float,
  style=listingcode,
  caption={urQL tests},
  label={lst:urql}
]
::
:: fail on dup rows
++  test-fail-insert-dup-rows
  =|  run=@ud
  =/  my-insert
    "INSERT INTO db1..my-table (col1, col2, col3)  ".
    "VALUES ('cord',~zod,20) ('Default',Default, 0)"
  %+  expect-fail-message
    'cannot add duplicate key:'
  |.  %-  process-cmds
      :+  gen3-dbs  :: <- one key already exists
        (bowl [run ~2031.1.1])
      (parse:parse(default-database 'db1') my-insert)
\end{lstlisting}
